Nothing lasts forever -- only item administration: An Item Response Theory algorithm to shorten tests

### 2025-02-28 sis2025/LaTex+Package/styles/paper-sis-template.tex
last update :)

\begin{document}
\mainmatter              % start of a contribution
%
\title{Nothing lasts forever -- only item administration: An Item Response Theory algorithm to shorten tests}
%
\titlerunning{IRT-based algorithms}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Ottavia M. Epifania\inst{1} and Livio Finos\inst{2}}
%
\authorrunning{Epifania \& Finos} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Ottavia M. Epifania, Livio Finos}
%
\institute{Department of Psychology and Cognitive Science, University of Trento, IT\\
\email{ottavia.epifania@unitn.it}
\and
{\color{red}Department of Statistics, University of Padova, IT} \livio{devi vedere se è giusto} \\}

@@

\begin{abstract}
\ottavia{da rivedere}
Although a larger number of test items improves measurement validity, the effect of respondents’ fatigue on the response quality should be acknowledged for developing reliable measurement tools.  This contribution presents an item response theory-based algorithm (denoted as Léon) able to shorten existing tests by concurrently accounting for the measurement precision of the abbreviated test and the tiredness of the respondents, which is here conceptualized as the probability of observing careless errors as the number of administered items increases. A simulation study compares the performance of Léon of approximating the measurement precision that would be obtained from the full-length test without the effect of the tiredness against that of another algorithm that does not account for the tiredness of the respondents. Although on average the two algorithms select the same number of items, Léon provides a better approximation to the measurement precision of the full-length test than the other algorithm.  
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{Item response theory, careless error, information functions, short test forms}
\end{abstract}
%
\section{Introduction}
%
As a general rule of thumb, the higher the number of items in a test, the better the measurement in terms of validity and reliability. However, there is a trade-off between the number of administered items and the response quality. As such, the trade-off between the number of administered items and the tiredness of the respondents \ottavia{(i.e., response fatigue)} should be kept in mind to obtain reliable and precise measurement tools. Item Response Theory (IRT, see, e.g., \cite{baker}) provides an ideal framework for shortening existing tests given the detailed information that they provide of the measurement precision of each item with respect to different levels of the latent trait (see e.g., \cite{pauci}). 
In this contribution, we present a new IRT-based algorithm for developing short test forms (STFs) from existing tests, denoted as Léon. This algorithm considers the response fatigue during the item selection process, such that it attempts at minimizing the number of selected items to maximize the measurement precision (as expressed by the test information function, TIF) of the STF.
%The ability of Léon of developing STFs able to approximate the TIF that would be obtained by administering all the items under the assumption that respondents would never get tired is investigated in a simulation study. 
%Specifically, Léon's ability is compared against that of another algorithm for developing STFs, which does not account for the tiredness of the respondents. 
The response fatigue has been here conceptualized as the careless error related to the item rank during the administration.

@@

\begin{equation}\label{eq:4pl}
	P(x_{pi}= 1| \theta_p, b_i, a_i, c_i, d_i) = c_i + (d_i -c_i) + \dfrac{\exp[a_i(\theta_p - b_i)]}{1 + \exp[a_i(\theta_p - b_i)]},
\end{equation}
where $\theta_p$ is the latent trait level of person $p$, $b_i$ is the item location on the latent trait (i.e., difficulty parameter, the higher the value, the higher the difficulty of the item), and $a_i$ is the item ability to discriminate between respondents with different latent trait levels (i.e., discrimination parameter, the higher the value, the higher the discrimination ability of the item). Parameters $c_i$ and $d_i$ are the lower and upper asymptotes of the probability, which should naturally tend to 0 and 1 when $\theta \to - \infty$ and $\theta \to +\infty$, respectively. 
%However, there might instances where respondents with $\theta_p \to -\infty$ provide the correct response and respondents with $\theta_p \to + \infty$ do not provide the correct response.
%However, there might be instances where respondents with $\theta$ levels below the difficulty of the item, whom are hence expected not to respond correctly, might provide the correct response out of luck. 
The pseudoguessing parameter $c_i$ is the probability of observing a correct response on $i$ for $\theta \to - \infty$, such that the probability tends to $c_i$ instead of 0. 
%The same but inverse consideration applies when $\theta \to + \infty$. 
The $d_i$ parameter is the probability of observing a correct response on $i$ for $\theta \to + \infty$, such that this probability tends to $d_i$ instead of 1. The careless error probability is $1 - d_i$ (i.e., the probability of not endorsing the item given that the $\theta_p > b_i$).

@@

\ottavia{In this study, the response fatigue has been conceptualized as the careless error probability $1 - d_i$, which increases as the administration goes on. As $1 - d_i$ increases, the upper aymptote $d_i$ decreases.}
This probability is instrinsically related to the item rank in the test, hence the careless error parameter in Equation (\ref{eq:4pl}), which is usually a property of the item ($d_i$), becomes a property of the item rank $r = \{1, \ldots, R\}$ in the test, $d_r$, with the constraint $d_{r-1} > d_r$.
\normalcolor

@@

\begin{equation}\label{eq:iif}
	\text{IIF}_{i}(r) = \dfrac{a^2[P(\theta)-c_i]^2[d_r - P(\theta)]^2}{(d_{r}-c_i)^2 P(\theta)Q(\theta)}, 
\end{equation}
whereas in the typical formalization $d_r = d_i$.
%the $d$ parameter in Equation (\ref{eq:iif}) is a property related to its order of presentation in the test $r$, and it is hence denoted as $d_r$ with $r = \{1, \ldots, R\}$.
The $\text{IIF}_i$ is strongly influenced by the item location on the latent trait with respect to a specific $\theta_p$, its discriminativity, and its lucky guess and careless error. In absence of lucky guess and careless error, the IIF is maximum when  $\theta_p = b_i$, and it decreases as the distance between $b_i$ and $\theta_p$ increases. When the lucky guess and careless error are taken into account, the informativeness of the item decreases, the IIF is lower, and it maximum is shifted.

@@

Since the TIF increases as the number of test items increases, the comparison between $\text{TIF}_B$ and the TIF of the STFs is based on the mean TIF (i.e., the TIF divided by the number of items). In what follows, the mean TIF  is simply referred to as $\text{TIF}_{Q_x}$, with $x \in \{B, B', \text{Frank}, \text{\text{Léon}}\}$ and $Q$ is the items set in the tests $B$ or $B'$ or the items subset in the STFs generated by the algorithms.

\subsection{Frank}

@@

Frank selects the item whose IIF is best able to reduce the distance from the $\text{TIF}_B$ along the entire latent trait, as follows:

At $k = 0$: $\text{IIF}_i$, $\text{TIF}^0(\theta) = 0 \, \forall \theta$, $Q^0 = \emptyset$. For $k \geq 0$,

@@

	\end{itemize}
\end{enumerate}
At $k = 0$, the subset of items $Q^0$ is empty and the $\text{TIF}^0$ is 0 for all the $\theta$ levels. 
At each iteration $k$: (1.) a set of available items is generated as the items in the item bank that have not been included in the STF yet, $A^k = B \setminus Q^k$; (2.)
An average provisional TIF, $\text{pTIF}$, is computed by adding the IIF of each of the items in the set of the available items $A^k$, one at the time, to the TIF obtained from the items in $Q^k$ (The denominator is obtained by adding 1 to the cardinality of $Q^k$); (3.) Among all the items in $A^k$, the one that allows for minimizing the distance between $\text{pTIF}$ and $\text{TIF}_B$ is included in $i^*$; (4.) 
The termination criterion is tested. 
If the distance between the $\text{TIF}_B$ and the $\text{pTIF}_{i^*}$ is greater than or equal to the distance between the $\text{TIF}_B$ and the $\text{TIF}^k$ (i.e., the TIF obtained from the items in the subset $Q^k$, without item $i^*$) (TRUE), then the item $i^*$ does not contribute in the reduction of the distance from the $\text{TIF}_B$, the algorithm stops, and the final item selection is the one without the item in $i^*$, $Q_{frank} = Q^k$. Conversely (FALSE), the item in $i^*$ does contribute in the reduction of the distance from the $\text{TIF}_B$, hence it is included in the set of items and a new iteration starts, $Q^{k+1} = Q^k \cup \{i^*\}$.

\color{blue}
\subsection{Léon}
Differently from Frank, at step (2.) Léon computes the $\text{IIF}_i$ by considering the $d_r$ related to the number of items included in the STF up to that point.  
\normalcolor

@@

The procedure is replicated 100 times. At each replication, a test $B$ of 50 items with difficulty ($b_i \sim \mathcal{U}(-3, 3)$) discrimination ($a_i \sim \mathcal{U}(.90, 2)$) parameters drawn from uniform distributions is generated. Lucky 
guess and careless error parameters are constant, $c_i = 0$ and $d_i = 1$, $\forall i \in B$.
The $\text{TIF}_B$ (i.e., the TIF-target) is obtained as the average TIF  from the items in $B$, and describes the measurement precision that would be obtained if respondents were administered with all the items without getting tired.

At each replication, a new test, $B'$, is generated, where $b_i' = b_i$, $a_i' = a_i$, $c_i = 0$, $\forall i \in B'$, and $d_r = \exp(-\lambda r_i)$, with $\lambda = 0.01$ and $r = \{0, \ldots, ||B|| -1\}$.

At each replication, Frank and Léon generate a STF for approximating the $\text{TIF}_B$. An average TIF is computed considering the item in $B'$, denoted as $\text{TIF}_{B'}$.
Although Frank grounds the item selection on the set $B$, the final TIF is computed by considering the vector $d_r$ associated to the rank of the items in $Q_{\text{Frank}}$.
%La differenza è che Frank calcola tutte le iif degli tiem all'inizio, consdierando la careless error legata alla posizione dell'item così come viene generata inizialmente dalla funzione. 
%Léon invece calcola la iif ad ogni itereazione (per questo è più lento) mettendo dentro la careless error che è legata alla posizione dell'item. Nel senso, se Léon per ora ha selezionato 3 item e deve valutare se metterne un quarto, le iif degli item che sono rimasti verrà calcolata considerando la careless error in $B'$ dell'item somministrato per quarto.
\normalcolor

@@

The average distance from $\text{TIF}_B$ of the STFs generated by Frank and Léon at each iteration, as well as of the test obtained from $B'$ has been considered as a criterion for evaluating the ability of Léon of generating informative STFs while accounting for the tiredness of the respondents, $\Delta_x = |\text{TIF}_B - \text{TIF}_{Q_x}|$, with $x \in \{B', \text{Frank}, \text{Léon}\}$.  Trivially, when the $\text{TIF}_{B'}$ is considered, $Q = B'$.

To better understand whether acknowledging the tiredness of the respondents influences the number of items included in $Q$, the cardinalities of the STFs generated by Frank and Léon, $||Q_{\text{Frank}}||$ and $||Q_{\text{Léon}}||$, respectively, have been compared.

@@

On average, Léon and Frank included the same number of items in the STF, $M_{\text{Léon}} = 6.75 \pm 3.06$, $M_{\text{Frank}} = 6.31 \pm 2.40$, $t = 1.13$, $df = 187.59$, $p = .26$, suggesting that accounting for the tiredness of the respondents does not influence the number of items included in the STF.
%\color{red}
%Given that leon account for the tiredness of the respondents during the item selection, it also considers the measurement precision of the STF given that the respondents do get tired during the administration, while Frank does not. As such, while preventing the over inclusion of items in the STF, Léon is also able to account for the measurement precision, in the attempt of balancing the number of items with the measurement precision.
%\normalcolor

Figure \ref{fig:points-alogirthms} illustrates the distributions of the distances from $\text{TIF}_B$ ($y$-axis) of $Q_x$ ($x$-axis).

>>>>

\begin{document}
\mainmatter              % start of a contribution
%
\title{Nothing lasts forever -- only item administration: An Item Response Theory algorithm to shorten tests}
%
\titlerunning{IRT-based algorithms}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Ottavia M. Epifania\inst{1} and Livio Finos\inst{2}}
%
\authorrunning{Epifania \& Finos} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Ottavia M. Epifania, Livio Finos}
%
\institute{Department of Psychology and Cognitive Science, University of Trento, IT\\
\email{ottavia.epifania@unitn.it}
\and
 Department of Statistical Sciences, University of Padova, IT \\}

@@

\begin{abstract} 
	\ottavia{The responses to the items composing psychometrics tests are often used to estimate the respondents' latent trait levels.}
Although a larger number of items improves measurement validity, the effect of respondents’ fatigue on the response quality should be acknowledged for developing reliable measurement tools.  This contribution presents an item response theory-based algorithm (denoted as Léon) able to shorten existing tests by concurrently accounting for the measurement precision of the abbreviated test and the tiredness of the respondents. A simulation study compares the performance of Léon of approximating the measurement precision that would be obtained from the full-length test without the effect of the tiredness against that of another algorithm that does not account for the tiredness of the respondents. Although, on average, the two algorithms select the same number of items, Léon provides a better approximation to the measurement precision of the full-length test than the other algorithm.  
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{Item response theory, careless error, information functions, short test forms}
\end{abstract}
%
\section{Introduction}
%
As a general rule of thumb, the higher the number of items in a test, the better the measurement in terms of validity and reliability. However, there is a trade-off between the number of administered items and the response quality. As such, the trade-off between the number of administered items and the tiredness of the respondents \ottavia{(i.e., response fatigue)} should be kept in mind to obtain reliable and precise measurement tools. Item Response Theory (IRT, see, e.g., \cite{baker}) provides an ideal framework for shortening existing tests given the detailed information that they provide of the measurement precision of each item with respect to different levels of the latent trait (see e.g., \cite{pauci}). 
In this contribution, we present a new IRT-based algorithm for developing short test forms (STFs) from existing tests, denoted as Léon. This algorithm considers the response fatigue during the item selection process, such that it attempts at minimizing the number of selected items to maximize the measurement precision (as expressed by the test information function, TIF) of the STF.
%The ability of Léon of developing STFs able to approximate the TIF that would be obtained by administering all the items under the assumption that respondents would never get tired is investigated in a simulation study. 
%Specifically, Léon's ability is compared against that of another algorithm for developing STFs, which does not account for the tiredness of the respondents. 
%The response fatigue has been here conceptualized as the careless error related to the item rank during the administration.

@@

\begin{equation}\label{eq:4pl}
	P(x_{pi}= 1| \theta_p, b_i, a_i, c_i, d_i) = c_i + (d_i -c_i) + \dfrac{\exp[a_i(\theta_p - b_i)]}{1 + \exp[a_i(\theta_p - b_i)]},
\end{equation}
where $\theta_p$ is the latent trait level of person $p$, $b_i$ is the item location on the latent trait (i.e., difficulty parameter, the higher the value, the higher the difficulty of the item), and $a_i$ is the item ability to discriminate between respondents with different latent trait levels (i.e., discrimination parameter, the higher the value, the higher the discrimination ability of the item). Parameters $c_i$ and $d_i$ are the lower and upper asymptotes of the probability, which should naturally tend to 0 and 1 when $\theta \to - \infty$ and $\theta \to +\infty$, respectively. 
%However, there might instances where respondents with $\theta_p \to -\infty$ provide the correct response and respondents with $\theta_p \to + \infty$ do not provide the correct response.
%However, there might be instances where respondents with $\theta$ levels below the difficulty of the item, whom are hence expected not to respond correctly, might provide the correct response out of luck. 
The pseudoguessing parameter $c_i$ is the probability of observing a correct response on $i$ for $\theta \to - \infty$, such that the probability tends to $c_i$ instead of 0. 
%The same but inverse consideration applies when $\theta \to + \infty$. 
The $d_i$ parameter is the probability of observing a correct response on $i$ for $\theta \to + \infty$, such that this probability tends to $d_i$ instead of 1. The careless error probability is $1 - d_i$ (i.e., the probability of not endorsing the item given that the $\theta_p > b_i$).
\ottavia{Item parameters are summarized in vector $\varphi = (b_i, a_i, c_i, d_r)$.}

@@

\ottavia{In this study, the response fatigue has been conceptualized as the careless error probability $1 - d_i$, which increases as the administration goes on. As $1 - d_i$ increases, the upper aymptote $d_i$ decreases.}
This probability is instrinsically related to the item rank in the test, hence the careless error parameter in Equation (\ref{eq:4pl}), which is usually a property of the item ($d_i$), becomes a property of the item rank $r = \{1, \ldots, R\}$ in the test, $d_r$, with the constraint $d_{r-1} > d_r$. \ottavia{per come l'ho scritto non uò essere maggiore uguale. credo che a livello teorico non abbia senso}
\normalcolor

@@

\begin{equation}\label{eq:iif}
	\text{IIF}_{i}(r) = \dfrac{a^2[P(x_{pi}=1|\theta, \varphi)-c_i]^2[d_r - P(x_{pi}=1|\theta, \varphi)]^2}{(d_{r}-c_i)^2 P(x_{pi}=1|\theta, \varphi)P(x_{pi}=0|\theta, \varphi)}, 
\end{equation}
whereas in the typical formalization $d_r = d_i$.
%the $d$ parameter in Equation (\ref{eq:iif}) is a property related to its order of presentation in the test $r$, and it is hence denoted as $d_r$ with $r = \{1, \ldots, R\}$.
The $\text{IIF}_i$ is strongly influenced by the item location on the latent trait with respect to a specific $\theta_p$, its discriminativity, and its lucky guess and careless error. In absence of lucky guess and careless error (i.e., $c = 0$, $d = 1$), the IIF is maximum when  $\theta_p = b_i$, and it decreases as the distance between $b_i$ and $\theta_p$ increases. When the lucky guess and careless error are taken into account, the informativeness of the item decreases, the IIF is lower, and it maximum is shifted.

@@

Since the TIF increases as the number of test items increases, the comparison between $\text{TIF}_B$ and the TIF of the STFs is based on the mean TIF (i.e., the TIF divided by the number of items). 
\ottavia{In what follows, the mean TIF  is referred to as $\text{TIF}_{x}$, with $x \in \{B, B', \text{Frank}, \text{\text{Léon}}\}$, $Q_x \subset B$ is the item subsets selected by Frank and Léon, and $||X||$ denotes the cardinality of set $X$.}

@@

Frank selects the item whose IIF is best able to reduce the distance from the $\text{TIF}_B$ along the entire latent trait, without considering the respondents' fatigue, as follows:

At $k = 0$: $\text{TIF}^0(\theta) = 0 \, \forall \theta$, $Q^0 = \emptyset$. For $k \geq 0$,

@@

	\end{itemize}
\end{enumerate}
At $k = 0$, the subset of items $Q^0$ is empty and the $\text{TIF}^0$ is 0 for all the $\theta$ levels. 
At each iteration $k$: (1.) a set of available items is generated as the items in the item bank that have not been included in the STF yet, $A^k = B \setminus Q^k$; (2.)
An average provisional TIF, $\text{pTIF}$, is computed by adding the IIF of each of the items in the set of the available items $A^k$, one at the time, to the TIF obtained from the items in $Q^k$ (the denominator is obtained by adding 1 to the cardinality of $Q^k$); (3.) Among all the items in $A^k$, the one that allows for minimizing the distance between $\text{pTIF}$ and $\text{TIF}_B$ is included in $i^*$; (4.) 
The termination criterion is tested. 
If the distance between the $\text{TIF}_B$ and the $\text{pTIF}_{i^*}$ is greater than or equal to the distance between the $\text{TIF}_B$ and the $\text{TIF}^k$ (i.e., the TIF obtained from the items in the subset $Q^k$, without item $i^*$) (TRUE), then the item $i^*$ does not contribute in the reduction of the distance from the $\text{TIF}_B$, the algorithm stops, and the final item selection is the one without the item in $i^*$, $Q_{frank} = Q^k$. Conversely (FALSE), the item in $i^*$ does contribute in the reduction of the distance from the $\text{TIF}_B$, hence it is included in the set of items and a new iteration starts, $Q^{k+1} = Q^k \cup \{i^*\}$.

To account for the respondents' fatigue, at step (2.) Léon computes the $\text{IIF}_i$ by considering the $d_r$ related to the number of items included in the STF up to that point.  
\normalcolor

@@

The following procedure is replicated 100 times. At each replication, a test $B$ of 50 items with difficulty ($b_i \sim \mathcal{U}(-3, 3)$) discrimination ($a_i \sim \mathcal{U}(.90, 2)$) parameters drawn from uniform distributions is generated. Lucky 
guess and careless error parameters are constant, $c_i = 0$ and $d_i = 1$, $\forall i \in B$.
The $\text{TIF}_B$ (i.e., the TIF-target) is obtained as the average TIF  from the items in $B$, and describes the measurement precision that would be obtained if respondents were administered with all the items without getting tired.

At each replication, a new test, $B'$, is generated, where $b_i' = b_i$, $a_i' = a_i$, $c_i = 0$, $\forall i \in B'$, and $d_r = \exp(-\lambda r)$, with $\lambda = 0.01$ and $r = \{0, \ldots, ||B|| -1\}$, the $\text{TIF}_B$ is computed, and Frank and Léon generate a STF for approximating it. 
An average TIF is computed considering the item in $B'$ ($\text{TIF}_{B'}$), as well as the TIFs from the STFs generated by Frank and Léon, $\text{TIF}_{\text{Frank}}$ and $\text{TIF}_{\text{Léon}}$, respectively.
Although Frank grounds the item selection on the set $B$, the final TIF is computed by considering the vector $d_r$ associated to the rank of the items in $Q_{\text{Frank}}$.
%La differenza è che Frank calcola tutte le iif degli tiem all'inizio, consdierando la careless error legata alla posizione dell'item così come viene generata inizialmente dalla funzione. 
%Léon invece calcola la iif ad ogni itereazione (per questo è più lento) mettendo dentro la careless error che è legata alla posizione dell'item. Nel senso, se Léon per ora ha selezionato 3 item e deve valutare se metterne un quarto, le iif degli item che sono rimasti verrà calcolata considerando la careless error in $B'$ dell'item somministrato per quarto.
\normalcolor

@@

The average distance between $\text{TIF}_B$ and $\text{TIF}_{\text{Frank}}$ and $\text{TIF}_B$ and $\text{TIF}_{\text{Léon}}$ at each iteration, as well as the  distance between $\text{TIF}_B$ and $\text{TIF}_{B'}$, have been considered as a criterion for evaluating the ability of Léon of generating informative STFs while accounting for the tiredness of the respondents, $\Delta_x = |\text{TIF}_B - \text{TIF}_{x}|$, with $x \in \{B', \text{Frank}, \text{Léon}\}$.  
To better understand whether acknowledging the tiredness of the respondents influences the number of items included in $Q$, the cardinalities of the STFs generated by Frank and Léon, $||Q_{\text{Frank}}||$ and $||Q_{\text{Léon}}||$, respectively, have been compared.

@@

On average, Léon and Frank included the same number of items in the STF, $M_{\text{Léon}} = 6.75 \pm 3.06$, $M_{\text{Frank}} = 6.31 \pm 2.40$, suggesting that accounting for the tiredness of the respondents does not influence the number of items included in the STF.
%\color{red}
%Given that leon account for the tiredness of the respondents during the item selection, it also considers the measurement precision of the STF given that the respondents do get tired during the administration, while Frank does not. As such, while preventing the over inclusion of items in the STF, Léon is also able to account for the measurement precision, in the attempt of balancing the number of items with the measurement precision.
%\normalcolor

Figure \ref{fig:points-alogirthms} illustrates the distributions of the distances from $\text{TIF}_B$ ($y$-axis) of $B'$, Frank, and Léon ($x$-axis).
